Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-05 paper/paper.tex
Added package makecell
correctness of liveness analysis

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
%\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage [plain,noend,noline,boxed]{algorithm2e}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
%\usepackage{program}
%\usepackage{algorithmic}
\usepackage[boxed]{algorithm2e}

@@

From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
that the results of liveness analysis of any program can be modeled by
the kind  of grammar described above. The  following proposition shows
that the converse also holds.
\begin{proposition}
Given    a     grammar    $G$    of    the     form    described    in
Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
program $p$  with program points  $\pi_i$ and variables $x$  such that
the liveness  analysis of $p$  is same as  $G$ except for a  change in
non-terminal names which now have the form \var{\Lanv{i}{x}}.
\end{proposition}
We now show that the decision problem required at garbage collection
time is undecidable.

@@

\comment{Proof:}

@@

\section{The Garbage Collection Scheme}
% GC assumptions
\cred{Introduction}

@@

\begin{figure*}[t!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          &\makecell{ $\rho, (\rho', x, e)\!:\!S, H, \kappa$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$ }    &  \sc{const}
\\
\hline
          & \makecell{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)$ \\ $\rightsquigarrow  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e)\!:\!S, H, (\CAR~x)$ \\ $ \rightsquigarrow \rho', S, H[\rho(z) := d_1], e$}      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S, H, (\CAR~x)$ \\ $\rightsquigarrow \rho', (\rho, x, (\CAR~x))\!:\!S, H, s$}      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \makecell{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)$ \\  $\rightsquigarrow \rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S, H, (+~x~y)$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y))\!:\!S, H, s$}      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho, S, H, (+~x~y)$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y))\!:\!S, H, s$}      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell{$\rho, S, H, (f~\myvec{x})$ \\ $\rightsquigarrow [\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$}      &
\sc{funcall} \\
\hline
$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\SIF~x~e_1~e_2), ep, \sigma$ \\ $\rightsquigarrow \rho, S, H,  e_1, ep, \sigma$} & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\SIF~x~e_1~e_2), ep, \sigma$ \\   $\rightsquigarrow
\rho, S, H,  e_2, ep, \sigma$} & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,  (\SIF~x~e_1~e_2), ep , \sigma$ \\ $\rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2), ep, \sigma)\!:\!S, H,  s, ep_{if}, \sigma$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho, (\rho', z, e, ep, \sigma)\!:\!S, H, (\SRETURN~x), ep_{return}, \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e, ep, \sigma$} &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H, (\SRETURN~x), ep, \sigma$ \\ $
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x), ep, \sigma)\!:\!S, H,  s, ep_{return}, \sigma$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics for the language.\label{fig:minefield-semantics}}
\end{center}
\end{figure*}

>>>>

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
%\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage [plain,noend,noline,boxed]{algorithm2e}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
%\usepackage{program}
%\usepackage{algorithmic}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}

@@

From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
that the results of liveness analysis of any program can be modeled by
the kind  of grammar described above. The  following proposition shows
that the converse also holds.
\begin{proposition}
Given    a     grammar    $G$    of    the     form    described    in
Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
program $p$  with program points  $\pi_i$ and variables $x$  such that
the liveness  analysis of $p$  is same as  $G$ except for a  change in
non-terminal names which now have the form \var{\Lanv{i}{x}}.
\end{proposition}
We now show that the decision problem required at garbage collection
time is undecidable.
\newcommand{\state}{\ensuremath{\mathsf{S}}}
\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}

@@

{\em Proof:}  Given a Turing machine  $TM$ and an input  $w$, we shall
construct an unrestricted  grammar $G$ of the kind  described above, so
that $TM$ will halt on $w$ if and only if $\epsilon \in \mathscr{L}(G)$.

We  shall  represent  the  configuration  of  $TM$  at  any  point  as
$w_L(\state,c)w_R$, where $w_L$ is the string on the left of the head,
$w_R$ is the string on the right  of the head, $c$ is the symbol under
the head and \state\ is the current state of the machine. For each
$(\state,c)$ combination of state and symbol, the grammar
will contain a nonterminal \nont{}{c}. Each step of the machine will
correspond to step of derivation based on the grammar.  In addition, we shall maintain
the following invariant between the configuration of $TM$ and the
current sentential form:

\begin{quote}
  If the configuration of the machine is $w_L(\state,c)w_R$, then the
  sentential form will be $\overline{w}_L\nont{}{c}\,w_R$, where
  $\overline{w}_L$ is the same as $w_L$ but with each symbol $d$
  replaced with $\overline{d}$.
\end{quote}

Assume that the TM starts in a state $S_{init}$ with a tape $cw$ and the
head is positioned on the symbol $c$. Then the start non-terminal  of the
grammar $\nont{}{}$ is associated with the production $\nont{}{} \rightarrow \nont{init}{c}w$ and the
sentential form corresponding to the initial configuration is
$\nont{init}{c}w$. Further each transition of the turing machine is modeled by
productions as follows:

\begin{enumerate}
\item A transition $(S_i, c) \rightarrow (S_j,c',L)$ by two productions
  $\nont{i}{c} \rightarrow \acar \nont{j}{\acar}c'$ and $\nont{i}{c}
  \rightarrow \acdr \nont{j}{\acdr}c'$. 
\item A transition $(S_i, c) \rightarrow (S_j,c',R)$ by two productions
  $\nont{i}{c} \rightarrow c \nont{j}{\acar}\bcar$ and $\nont{i}{c}
  \rightarrow c \nont{j}{\acdr}\bcdr$. 
\end{enumerate}
The idea behind these productions is as follows: Assume that the
current sentential form is
\ldots\bcar\bcdr\nont{i}{\acar}\acar\acar\ldots and Starting from $(S_i,
c)$ if the head writes $c'$ and moves to the left, then the symbol
$c'$ is exposed to the right of the tape head. However, we do not know
which symbol that is under the head after the movement. Therefore we
generate two productions assuming that the symbols are \acar\ and
\bcar. If the guess is right, then to remove the symbol to the left in
the sentential form

@@

\section{The Garbage Collection Scheme}

@@

\begin{figure*}[t!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          &\makecell{ $\rho, (\rho', x, e, \{\psi_{i}\}, \sigma')\!:\!S, H, \kappa, \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e, \sigma'$ }   &  \sc{const}
\\
\hline
          & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma')\!:\!S, H, (\CONS~x~y), \sigma$ \\ $\rightsquigarrow  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e, \sigma'$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma' )\!:\!S, H, (\CAR~x), \sigma$ \\ $ \rightsquigarrow \rho', S, H[\rho(z) := d_1], e, \sigma'$}      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S, H, (\CAR~x), \sigma$ \\ $\rightsquigarrow \rho', (\rho, x, (\CAR~x), \{\psi_{i}\}, \sigma)\!:\!S, H, s,\bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{x} $}      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma')\!:\!S, H, (+~x~y), \sigma$ \\  $\rightsquigarrow \rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e, \sigma'$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S, H, (+~x~y), \sigma$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y), \{\psi_{i}\}, \sigma)\!:\!S, H, s, \bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{x}$}      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho, S, H, (+~x~y), \sigma$ \\ $\rightsquigarrow \rho', (\rho, y, (+~x~y), \{\psi_{i}\}, \sigma)\!:\!S, H, s, \bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{y} $}      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell{$\rho, S, H, (f~\myvec{x}), \sigma$ \\ $\rightsquigarrow [\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}, \sigma$}      &
\sc{funcall} \\
\hline
$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2), ep, \sigma$ \\ $\rightsquigarrow \rho, S, H,  e_1, ep, \sigma$} & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2), ep, \sigma$ \\   $\rightsquigarrow
\rho, S, H,  e_2, ep, \sigma$} & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,  (\pi:\SIF~\psi:x~e_1~e_2), \sigma$ \\ $\rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2),  \pmap{}(\pi), \sigma)\!:\!S, H,  s, \mmap{}{\psi}{x}$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho, (\rho', z, e, ep, \sigma')\!:\!S, H, (\SRETURN~x), \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e, \sigma'$} &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H, (\psi:\SRETURN~x), \sigma$ \\ $
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x), \{\psi\}, \sigma)\!:\!S, H,  s, \mmap{}{\psi}{x}$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics for the language.\label{fig:minefield-semantics}}
\end{center}
\end{figure*}
